---
title: "Homework before session 3"
---

```{r, include=FALSE}
source("setup.R")
knitr::opts_chunk$set(fig.path = paste0(knitr::opts_chunk$get("fig.path"),
                                        "07-"))
```


For our homework exercises, we will use a new dataset from the [_Gapminder Foundation_](https://www.gapminder.org/), 
which gives access to global data as well as many tools to help explore it.

We will use data relating to socio-economic statistics for 2010. 
The columns in our data file are:

| Column                       | Description                                                                                                                     |
|-----------------------------|---------------------------------------------------------------------------------------------------------------------------------|
| country                     | country name                                                                                                                    |
| world_region                | 6 world regions                                                                                                                 |
| year                        | year that each datapoint refers to                                                                                               |
| children_per_woman          | total fertility rate                                                                                                            |
| life_expectancy             | average number of years a newborn child would live if current mortality patterns were to stay the same                          |
| income_per_person           | gross domestic product per person adjusted for differences in purchasing power                                                  |
| is_oecd       | Whether a country belongs to the “OECD” (`TRUE`) or not (`FALSE`)                                                                                                       |
| income_groups               | categorical classification of income groups                                                                                     |
| population             | total number of a country's population                                                                                                 |
| main_religion               | religion of the majority of population in 2008                                                                                  |
| child_mortality             | death of children under 5 years old per 1000 births                                                                             |
| life_expectancy_female      | life expectancy at birth, females                                                                                               |
| life_expectancy_male        | life expectancy at birth, males                                                                                                 |


## Task 1 - setting up your project

:::note
**Tip:** this exercise builds on the skills gained in [Getting Started](00-getting-started.html).
:::

1. On your computer, create a new directory for this project called `gapminder-dataviz`.
1. Create directories that you feel are important, including one called `data_raw` for saving the raw data.
1. Download the [gapminder socio-economic dataset 2010](https://github.com/tavareshugo/r-eda-gapminder/raw/gh-pages/_episodes_rmd/data/raw/gapminder2010_socioeconomic.csv)
and save it in your project's data folder. 
(<kbd>right-click</kbd> the file link and choose "Save link as...")
1. Create a new R Project on the `gapminder-dataviz` directory you just created.


<details><summary>Answer</summary>

We create a new directory as well as sub-directories, shown here schematically:

```
gapminder-dataviz
 |_ data_raw
 |_ data_processed
 |_ fig_output
 |_ scripts
```

We use `data_raw` to save the data file that we download with the link provided. 

Finally, we create an R project on this directory:

1. Start RStudio.
2. Under the <kbd>File</kbd> menu, click on <kbd>New Project</kbd>. Choose <kbd>Existing Directory</kbd>.
3. Hit the <kbd>Browse...</kbd> button to go to the folder named `gapminder-dataviz` that we've just created.
4. Click on <kbd>Create Project</kbd>.

RStudio should refresh itself and then indicate that the _working directory_ has been set to the new folder. 
For example, you can run the command `getwd()` on the console to confirm that this is the case.
 
</details>


## Task 2 - importing data

As with any dataset you must first understand its content and formatting.
Understanding what data you have will help you decide what story you can learn from the data and how best to present it

Create a new script to analyse these data and call it `01-gapminder_exploration.R`. 
Then, populate it with code to achieve the following:

1. Read the `gapminder2010_socioeconomic.csv` file into a `data.frame`/`tibble` object 
called `gapminder`. <details><summary>Hint</summary>Use the `read_csv()` function. Remember to first load the `tidyverse` package with `library(tidyverse)`.</details>
1. Make a basic check that your data import went well. Your data frame should have 
193 rows and 13 columns. <details><summary>Hint</summary>Use functions such as `nrow()`, `ncol()`, `summary()` and `str()` to check data integrity.</details>
1. What types of variables does this data have?
1. Are there any issues that you notice with these data? Note these down. 

<details><summary>Answer</summary>

We can read our data as follows:

```{r, warning=FALSE}
library(tidyverse)

gapminder <- read_csv("data_raw/gapminder2010_socioeconomic.csv")
```

To examine the contents of the data.frame we can use several functions, for example, to get the number of rows and columns:

```{r}
ncol(gapminder) # number of columns in the data.frame
nrow(gapminder) # number of rows in the data.frame
```

The `str()` function gives a more comprehensive view of the contents of the data.frame, including the number of rows and columns as well as the type of variable each columns was imported as by R:

```{r}
str(gapminder)
```

The summary function is also very useful as it gives a quick overview of the types of variables as well as average and quantiles for numeric data:

```{r}
summary(gapminder)
```

The types of variables we have are:

- nominal (categorical variable with no order): for example `country` and `world_region`
- ordinal (categorical variable with order): for example `income_groups`
- binary (categorical variable with two mutually exclusive values): `is_oecd`
- continuous (numerical variable with any values): for example `income_per_person` or `life_expectancy`
- discrete (numerical variable with limited values only): for example `year`

From the output of the summary function, we can notice a few issues with these data:

- Some `life_expectancy_male` values are invalid == -999.
- `life_expectancy_female` was imported as a character variable, but should be numeric.
- Some missing data - this is almost always expected in large datasets. 

If we look at the top few rows of the table:

```{r}
head(gapminder)
```

We can see that `life_expectancy_female` is showing the value "-" in the 4th row of data. 
Probably the person recording these data encoded missing values with the "-" symbol, but the `read_csv()` function did not recognise this as missing data. 
The default is to consider _empty_ cells as missing data and so we should correct this in the dataset to make sure **all missing values are encoded in the same way**.

There are a few other issues in the `main_religion` column, which were a little harder to detect. 
If we look at the unique values of this column, we will notice different spellings/formats for some of its values:

```{r}
# check unique values in a column
unique(gapminder$main_religion)
```

These types of spelling mistakes are very common and it's important to be aware that R would consider "muslim" and "Muslim" to be different words (due to the case-sensitivity). 

</details>


## Task 3 - plotting

1. Create a scatter plot for life expectancy (x) against income per person (y)
1. You will notice you get a warning message here - why do you think this is?

<details><summary>Answer</summary>

```{r}
ggplot(data = gapminder, 
       mapping = aes(x = life_expectancy, y = income_per_person)) +
  geom_point()
```

The warning message we get is because 6 of the rows in the data frame do not have life_expectancy information (they are `NA` missing data).

If it was a key variable for your analysis you might have wanted to remove those individuals with missing data. 
In this case, we don't mind having these missing data, so we can carry on with our analysis. 

Note that 'Warning messages' are simply that - a warning, not an error. 
They are very helpful and always worth reading.
  
</details>


## Task 4 - data manipulation

1. How many countries are there in "South Asia"? <details><summary>Hint</summary>Use the `filter()` function to subset the table to retain only rows where `world_region == "south_asia"`.</details>
1. Create a new column in the table called `income_total`, which is the product of population and income per person (i.e. the total average income of the country). <details><summary>Hint</summary>Use the `mutate()` function to create a new column.</details>

<details><summary>Answer</summary>

To identify countries in South Asia, we can use the following:

```{r, purl=TRUE, eval=TRUE}
gapminder %>% 
  filter(world_region == "south_asia")
```

From the output we can see this table has 8 rows, therefore 8 countries in this part of the world. 

To create the new column we can use the `mutate()` function, as such:

```{r}
gapminder %>% 
  mutate(income_total = population*income_per_person)
```

If we wanted to save this in our table, we need to update the object, using the `<-` assignment:

```{r}
gapminder <- gapminder %>% 
  mutate(income_total = population*income_per_person)
```

</details>
