---
title: Homework 12 August 2021
---


```{r, include=FALSE}
source("setup.R")
knitr::opts_chunk$set(fig.path = paste0(knitr::opts_chunk$get("fig.path"),
                                        "09-"))
```

These exercises continue from last week's, so you need to make sure you've read the gapminder dataset:

```{r, message=FALSE}
gapminder <- read_csv("data_raw/gapminder2010_socioeconomic.csv")
```



## Task 7

Let's take the scatterplot from our previous exercise and continue working on it:

```{r, warning=FALSE}
income_life_expectancy_plot <- ggplot(data = gapminder, 
                                      mapping = aes(x = life_expectancy, y = income_per_person)) +
  geom_point(aes(colour=world_region)) + 
  geom_smooth()
```


1. Using facets, split your scatter plot from the previous task by the variable `is_oecd`.
<details><summary>Hint</summary>Clicking the "Zoom" button on the bottom right "Plots" panel of RStudio to see your facetted graph much better (and any other graph for that matter!).</details>
1. Modify the previous code to facet your plot by `main_religion`. Do you identify any issues in this visualisation?
1. Finally, try facetting by `children_per_woman`. What do you think about this visualisation?

<details><summary>Answer</summary>

We can use the `facet_wrap()` function to achieve this:

```{r}
income_life_expectancy_plot  +
  facet_wrap(facets = vars(is_oecd))
```

To facet by a different variable, we can adapt the code above and just replace the variable we are facetting by:

```{r}
income_life_expectancy_plot  +
  facet_wrap(facets = vars(main_religion))
```

This is where we can clearly see how the spelling/formatting mistakes affect our data.
Variables are recognised as different if capitalisation is different e.g. "christian" vs "Christian".
There is also a more subtle typo between "eastern religions" (one space between words) and "eastern  religions" (two spaces between words). 
We turn to try and fix this problem in the next exercise. 

Finally, we were asked to facet by `children_per_woman`

```{r, warning=FALSE}
income_life_expectancy_plot +
  facet_wrap(facets = vars(children_per_woman))
```

This "works" (as in, the code gives no error). But it is very messy and not informative!
This is a good example of how NOT to present data - this would be a more suitable variable to use for "coloured by" as opposed to facet.

</details>


## Task 8

In previous exercises we have seen that the `main_religion` column contains several typos. 
The `stringr` package (part of `tidyverse`) is very useful to correct these kinds of mistakes. 
For example, the function `str_squish()` removes any whitespaces inside a string. 
While the function `str_to_lower()` converts words to lowercase. 

You have also seen how to use the `mutate()` function to _modify_ a column in your table. 
For example:

```{r, results='hide'}
gapminder %>% 
  mutate(income_total = population * income_per_person)
```

1. Use `mutate()`, `str_squish()` and `str_to_lower()` to fix the typos in the `main_religion` column in our table. 

<details><summary>Answer</summary>

We can pipe two `mutate()` commands, at each step modifying the column with one of the `str_*()` functions:

```{r}
gapminder <- gapminder %>% 
  mutate(main_religion = str_squish(main_religion)) %>% 
  mutate(main_religion = str_to_lower(main_religion))
```

Alternatively, we could have also done both tasks at once by "nesting" one function inside the other:

```{r, results='hide'}
gapminder %>% 
  mutate(main_religion = str_squish(str_to_lower(main_religion)))
```

We can check that the issue has been solved, for example by re-doing our plot facetted by this variable:

```{r}
# create new version of the base plot
income_life_expectancy_plot <- ggplot(gapminder, aes(life_expectancy, income_per_person)) +
  geom_point() +
  geom_smooth()

# plot with facets
income_life_expectancy_plot +
  facet_wrap(facets = vars(main_religion))
```

</details>



## Task 9

1. Split your scatter plot by rows and columns. Row = `is_oecd`, Columns = `main_religion`. 
1. Apart from the warning related to 6 missing values, we get other warnings. What do you think these are related to?

<details><summary>Answer</summary>

In this case, we want to use `facet_grid()`, which lets us split the plot by "rows" and "columns" on a grid:

```{r}
income_life_expectancy_plot  +
  facet_grid(rows = vars(is_oecd), cols = vars(main_religion))
```

The warnings we get are from `geom_smooth()`, which in some groups has such few points that it cannot make a smooth line (e.g. very few countries that are both part of the OECD and where the main religion falls in the category "eastern religions"). 

</details>


## Task 10

1. Save one of the plots (of your choice) to a file with appropriate file name. Choose the one you think best represents the data and think why.

<details><summary>Answer</summary>

We can use the `ggsave()` function to achieve this:

```{r, eval=FALSE}
ggsave(filename = "img_output/plot_income_lifeExpt_trend.png", 
       plot = income_life_expectancy_plot)
```

The extension we use in the filename is important, as it will determine the file format that the plot will be saved as. 
In this case we're saving it as a _PNG_ file. 
Another good option is to save as a _PDF_ file (`.pdf` extension).

We chose to just save the plot pre-facetting. 
This is because, sometimes, too many variables represented in a single plot confuses things and makes it extremely hard to interpret (and very easy to misinterpret!).

</details>
